---
title: "Movie_ratings_lab_2"
author: "Sneha Vasanth"
date: "January 19, 2017"
output: html_document
---
#1.Load the data.
```{r}
ratings <- read.csv("ratings.csv")
titles <- read.csv("movie.titles.csv")
```

a)	(2 pt) How many unique move genres are there? Note: simply calling unique() on the genre column is not the right answer. You will have to figure out how to count the individual genres. For example: the genre entry "Animation|Children|Drama|Musical" has 4 unique genres: "animation", "children", "drama", and "musical". Potentially useful functions are strsplit() and unlist().

```{r}

my_str <- strsplit(as.character(ratings$genre), "\\|")
my_unlist <- unlist(my_str)
my_table <- table(my_unlist)
my_table
dim(my_table)
```
There are 20 unique genre of movies.


b)	(2 pt) Of these unique genres what are the 5 most popular? (ie, which 5 the highest numbers of movies?)

```{r}
sorted_table <- sort(my_table, decreasing = TRUE)
head(sorted_table , n=5)
```

c)	(3 pt) What does the distribution of number of ratings per user look like? Are there "super raters" who review much more than others?

Yes, some of the raters review much more than others, as revealed in the histogram.

```{r}
library(dplyr)
count_rating_per_user <- 
  ratings %>%
  group_by(userId) %>%
  summarise(mycount = n())

hist(count_rating_per_user$mycount, xlab = "Number of ratings")
  
```

d)	(3 pt) Plot the number of movies rated by year of the most prolific rater. What does the trend look like?

The number of movies reviewed increase linearly over the years. 
 
```{r}
library(dplyr)
top_raters <- 
  ratings %>%
  group_by(userId) %>%
  summarise(my_count = n()) %>%
  arrange(desc(my_count))

head(top_raters, n=1)
```
Most prolific rater is userId 547 with 2391 ratings.

```{r}
ratings_by_year <- ratings %>%
  group_by(year) %>%
  filter(userId == 547) %>%
  summarise(my_count = n())

plot(ratings_by_year)
#hist(ratings_by_year$year, ratings_by_year$my_count)
#hist(ratings_by_year)
```


